A geometric framework for capturing high-resolution insights into conceptual knowledge and learning in classroom-like settings

### 2023-01-28 paper/main.tex
saving progress on edits

@@

\title{Text embedding models reveal high resolution insights into conceptual
knowledge from short multiple choice quizzes}

\author{Paxton C. Fitzpatrick\textsuperscript{1},
Andrew C. Heusser\textsuperscript{1, 2}, and Jeremy R.
Manning\textsuperscript{1, *}\\\textsuperscript{1}Dartmouth
College\\\textsuperscript{2}Akili Interactive\\\textsuperscript{*}Corresponding author:
jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract} We develop a mathematical framework, based on natural language
processing models, for tracking and characterizing the acquisition of
conceptual knowledge. Our approach embeds each concept in a high dimensional
representation space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
answered small sets of multiple choice quiz questions interleaved between
watching two course videos from the Khan Academy platform. We applied our
framework to the videos' transcripts, and to text of the quiz questions, to
quantify the content of each moment of video and each quiz question. We used
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings show how a
small set of quiz questions may be used to obtain rich and meaningful high
resolution insights into what each learner knows, and how their knowledge
changes over time as they learn.

@@

Suppose that a teacher had access to a complete ``map'' of everything their
student knew. Defining what such a map might even look like, let alone how it
might be constructed or filled in, is itself a non-trivial problem. But if a
teacher \textit{were} to gain access to such a map, how might that change their
ability to teach the student? Perhaps they might start by checking how well the
student knew the to-be-learned information already, or how much they knew about
related concepts. For some students, they could potentially optimize their
teaching efforts to maximize efficiency by focusing primarily on not-yet-known
content. For other students (or other content areas), it might be more
effective to optimize for direct connections between already-known content and
any new material. Observing how the student's knowledge was changing over time,
in response to their training, could also help to guide the teacher.

@@

The difference between ``understanding'' and ``memorizing,'' as framed by the
researchers in education, cognitive psychology, and cognitive
neuroscience~\citep{Kato40, Gall00, ScotEtal07, HallGree08, Macl05} has
profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
information about what the document is about, just as memorizing a passage
might be used to answer simple questions about the passage~\citep[e.g., whether
it might contain words related to furniture versus physics;][]{LandDuma97,
BleiEtal03, BleiLaff06}. However, modern natural language processing
models~\citep[e.g.,][]{MikoEtal13a, CerEtal18, BrowEtal20} also attempt to
capture the deeper meaning \textit{underlying} those atomic elements. These
models consider not only the co-occurrences of those elements within and across
documents, but also patterns in how those elements appear across different
scales (e.g., sentences, paragraphs, chapters, etc.), the temporal and
grammatical properties of the elements, and other high-level characteristics of
how they are used~\citep{Mann20, Mann21a}. According to these models, the deep
conceptual meaning of a document may be captured by a feature vector in a
high-dimensional representation space, where nearby vectors reflect
conceptually related documents. A model that succeeds at capturing an analog of
``understanding'' is able to assign nearby feature vectors to two conceptually
related documents, \textit{even when the words contained in those documents
have very little overlap}.

Given these insights, what form might the representation of the sum total of a
person's knowledge take? First, we might require a means of systematically
describing or representing the nearly infinite set of possible things a person
could know. Second, we might want to account for potential associations between
different concepts. For example, the concepts of ``fish'' and ``water'' might
be associated in the sense that fish live in water. Third, knowledge may have a
critical dependency structure, such that knowing about a particular concept
might require first knowing about a set of other concepts. For example,
understanding the concept of a fish swimming in water first requires
understanding what fish and water \textit{are}. Fourth, as we learn, our
``current state of knowledge'' should change accordingly. Learning new concepts
should both update our characterizations of ``what is known'' and should also
unlock any now-satisfied dependencies of that newly learned concept so that
they are ``tagged'' as available for future learning.

>>>>

% simple command for inline comments
\newcommand{\comment}[1]{}

@@

%\title{Text embedding models reveal high-resolution insights into conceptual
%knowledge and learning from short multiple-choice quizzes}
\title{A geometric framework for capturing high-resolution insights into
conceptual knowledge and learning in classroom-like settings}

\author{Paxton C. Fitzpatrick\textsuperscript{1},
Andrew C. Heusser\textsuperscript{1, 2}, and Jeremy R.
Manning\textsuperscript{1, *}\\\small{\textsuperscript{1}Department of Psychological and Brain Sciences}\\\small{Dartmouth College, Hanover, NH 03755, USA}\\\small{\textsuperscript{2}Akili Interactive Labs}\\\small{Boston, MA 02110, USA}\\\small{\textsuperscript{*}Corresponding author:
Jeremy.R.Manning@Dartmouth.edu}}

@@

\begin{abstract}\noindent We develop a mathematical framework, based on natural language
processing models, for tracking and characterizing the acquisition of
conceptual knowledge in real-world educational contexts. Our approach embeds
course content in a high-dimensional
conceptual space, where nearby coordinates reflect similar or related
concepts. We test our approach using behavioral data from participants who
watched two lecture videos from the Khan Academy platform, interleaved
between three short multiple-choice quizzes. We applied our
framework to the videos' transcripts and the text of the quiz questions to
quantify the conceptual content presented in each moment of video and knowledge probed by each quiz question. We used
these embeddings, along with participants' quiz responses, to track how the
learners' knowledge changed after watching each video. Our findings demonstrate how a
small set of quiz questions may be used to obtain rich and meaningful 
high-resolution insights into individual students' knowledge, and how it
changes over time as they learn.

@@

Suppose that a teacher had access to a complete ``map'' of everything a
student knew. Defining what such a map might even look like, let alone how it
might be constructed or filled in, is itself a non-trivial problem. But if a
teacher \textit{were} to gain access to such a map, how might it change their
ability to teach that student? Perhaps they might start by checking how well the
student knew the to-be-learned information already, or how much they knew about
related concepts. For some students, they could potentially optimize their
teaching efforts to maximize efficiency by focusing primarily on not-yet-known
content. For other students (or other content areas), it might be more
effective to optimize for direct connections between already known content and
new material. Observing how the student's knowledge changed over time,
in response to their teaching, could also help to guide the teacher towards
the most effective strategy for that individual student.

@@

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience~\citep[e.g.,][]{Kato40, Gall00, ScotEtal07, HallGree08, Macl05} has
profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
information about what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it~\citep[e.g., whether
it might contain words related to furniture versus physics;][]{LandDuma97,
BleiEtal03, BleiLaff06}. However, modern natural language processing
models~\citep[e.g.,][]{MikoEtal13a, CerEtal18, BrowEtal20} also attempt to
capture the deeper meaning \textit{underlying} those atomic elements. These
models consider not only the co-occurrences of those elements within and across
documents, but also patterns in how those elements appear across different
scales (e.g., sentences, paragraphs, chapters, etc.), the temporal and
grammatical properties of the elements, and other high-level characteristics of
how they are used~\citep{Mann20, Mann21a}. According to these models, the deep
conceptual meaning of a document may be captured by a feature vector in a
high-dimensional representation space, where nearby vectors reflect
conceptually related documents. A model that succeeds at capturing an analog of
``understanding'' is able to assign nearby feature vectors to two conceptually
related documents, \textit{even when the words contained in those documents
have very little overlap}.

Given these insights, what form might the representation of the sum total of a
person's knowledge take? First, we might require a means of systematically
describing or representing the nearly infinite set of possible things a person
could know. Second, we might want to account for potential associations between
different concepts. For example, the concepts of ``fish'' and ``water'' might
be associated in the sense that fish live in water. Third, knowledge may have a
critical dependency structure, such that knowing about a particular concept
might require first knowing about a set of other concepts. For example,
understanding the concept of a fish swimming in water first requires
understanding what fish and water \textit{are}. Fourth, as we learn, our
``current state of knowledge'' should change accordingly. Learning new concepts
should both update our characterizations of ``what is known'' and also
unlock any now-satisfied dependencies of those newly learned concepts so that
they are ``tagged'' as available for future learning.
% TODO: is the second half of this paragraph relevant to this paper??
